---
title: "2018 Bacterial Abundance"
author: "Christina Jimenez"
date: "10/28/2020"
output: github_document
---

# Introduction

This document will be an analysis of the 2018 bacterial abundance data.

# Load packages

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(patchwork)
```

# Import data

```{r}
excel_sheets("~/Desktop/GITHUB/144l_students/Input_Data/week3/144L_2018_BactAbund.xlsx")
```

```{r}
metadata <- read_excel("~/Desktop/GITHUB/144l_students/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Metadata")

glimpse(metadata)
```


```{r}
data <- read_excel("~/Desktop/GITHUB/144l_students/Input_Data/week3/144L_2018_BactAbund.xlsx", sheet = "Data")

glimpse(data)
```

```{r}
joined <- left_join(metadata, data)
glimpse(joined)  
```

# Prepare data

```{r}
cells <- joined %>% 
  mutate(Datetime = ymd_hm(Datetime),
         cells = Cells_ml * 1000) %>% 
  group_by(Experiment, Treatment, Bottle) %>%
  mutate(interv = interval(first(Datetime), Datetime),
         hours = interv/3600,
         days = hours/24) %>%
  ungroup() %>%
  select(Experiment:DNA_Sample, hours, days, cells) %>%
  drop_na(cells)

glimpse(cells)
```

# Plot growth curves

```{r fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
custom.colors <- c("Control" = "#00D3FFFF","Ash Leachate" = "#42FF00FF", "Mud Leachate" = "#FFF318FF", "Glucose_Nitrate_Phosphate" = "#4C00FFFF")
levels <- c("Control", "Ash Leachate", "Mud Leachate", "Glucose_Nitrate_Phosphate")

cells %>% 
  ggplot(aes(x = days, y = cells, group = interaction(Experiment, Treatment, Bottle))) +
  geom_errorbar(aes(ymin = cells, ymax = cells, color = factor(Treatment, levels = levels))) +
  geom_line(aes(fill = factor(Treatment, levels = levels)), size = .5) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 2, color = "black", shape = 21) +
  labs(x = "Days", y = expression(paste("Cells, L"^-1)), fill = "") + 
  guides(color = F) + 
  scale_color_manual(values = custom.colors) + 
  scale_fill_manual(values = custom.colors) + 
  theme_bw()
```

# Exponential phase of growth

```{r}
ln_cells <- cells %>%
  group_by(Experiment, Treatment, Bottle) %>%
  mutate(ln_cells =log(cells),
         diff_ln_cells = ln_cells - lag(ln_cells, default = first(ln_cells))) %>% 
  ungroup()
  
```

```{r fig.height=4, fig.width=12}
ln_cells %>% 
  mutate(dna = ifelse(DNA_Sample == T, "*", NA)) %>%
  ggplot(aes(x = days, y = diff_ln_cells, group = interaction(Experiment, Treatment, Bottle))) + 
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 0.5) +
  geom_point(aes(fill = factor("Treatment", levels = levels)), size = 2, color = "black", shape = 21) + 
  geom_text(aes(label = dna), size = 8, color = "#E41A1C") +
  labs(x = "Days", y = expression(paste("∆ln Cells, L"^-1)), fill = "") +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  facet_grid(Location~Bottle, scales = "free") +
  theme_bw()
```

- Control Bottle A \ ~0-1 d (T0-T1)
- Control Bottle B \ ~0-1 d (T0-T1)
- Ash Leachate Bottle C \ ~0-0.5 d (T0-T1)
- Ash Leachate Bottle D \ ~0-0.5 d (T0-T1)
- Mud Leachate Bottle E \ ~0-0.5 d (T0-T1)
- Mud Leachate Bottle F \ ~0-0.5 d (T0-T1)
- Glucose_Nitrate_Phosphate Bottle G \ ~0-1 d (T0-T1)
- Glucose_Nitrate_Phosphate Bottle H \ ~0-1 d (T0-T1)

# Growth rates, doubling times, and ∆cell abundances

```{r}
growth <- ln_cells %>% 
  mutate(exp_start = ifelse(Experiment == "144L_2018" & Bottle == "A", 0, NA),
         exp_start = ifelse(Experiment == "144L_2018" & Bottle %in% c("B", "C", "D", "E", "F", "G", "H"), 0, exp_start),
         exp_end = ifelse(Experiment == "144L_2018" & Bottle %in% c("A", "B", "C", "D", "E", "F", "G", "H"), 1)) %>%
  group_by(Experiment, Treatment, Bottle) %>% 
  mutate(ln_cells_exp_start = ifelse(Timepoint == exp_start, ln_cells, NA),
         ln_cells_exp_end = ifelse(Timepoint == exp_end, ln_cells, NA),
         cells_exp_start = ifelse(Timepoint == exp_start, cells, NA),
         cells_exp_end = ifelse(Timepoint == exp_end, cells, NA),
         days_exp_start = ifelse(Timepoint == exp_start, days, NA),
         days_exp_end = ifelse(Timepoint == exp_end, days, NA)) %>% 
  fill(ln_cells_exp_start:days_exp_end, .direction = "updown") %>% 
  mutate(mew = (ln_cells_exp_end - ln_cells_exp_start)/(days_exp_end - days_exp_start),
         doubling = log(2)/mew,
         delta_cells = cells_exp_end - first(cells)) %>% 
  ungroup()


glimpse(growth)
```

# Convert bacterial abundance & change in bacterial abundance to carbon units

```{r}
bactcarbon <- growth %>% 
  mutate(bc = cells *(2.5 * 10^-9),
         delta_bc = delta_cells * (2.5 *10^-9))

glimpse(bactcarbon)
```

# Calculate treatment averages

```{r}
averages <- bactcarbon %>% 
  group_by(Experiment, Treatment, Timepoint) %>% 
  mutate(ave_bc = mean(bc),
         sd_bc = sd(bc)) %>% 
  ungroup() %>% 
  group_by(Experiment, Treatment) %>% 
  mutate(ave_mew = mean (mew),
         sd_mew = sd(mew),
         ave_doubling = mean(doubling),
         sd_doubling = sd(doubling),
         ave_delta_cells = mean(delta_cells),
         sd_delta_cells = sd(delta_cells),
         ave_delta_bc = mean(delta_bc),
         sd_delta_bc = sd(delta_bc)) %>% 
  ungroup()
  

glimpse(averages)  
```

# Plot treatment averages

```{r fig.height=4, fig.width=6}
averages %>% 
  ggplot(aes(x = days, y = ave_bc), group = interaction(Experiment, Treatment)) +
  geom_errorbar(aes(ymin = ave_bc - sd_bc, ymax = ave_bc + sd_bc, color = factor(Treatment, levels = levels))) +
  geom_line(aes(color = factor(Treatment, levels = levels)), size = 0.5) +
  geom_point(aes(color = factor(Treatment, levels = levels)), color = "black", shape = 21, size = 2) +
  scale_color_manual(values = custom.colors) +
  scale_color_manual(values = custom.colors) +
  labs(x = "Days", y = expression("Bacterial Carbon, umol C L"^-1), fill = "", color = "") +
  guides(color = F) +
  theme_bw()
```

## Barplots

```{r}
bar.data <- averages %>% 
  select(Location, Treatment, ave_mew:sd_delta_bc) %>% 
  distinct()
```


```{r}
mew <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y =ave_mew), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_mew - sd_mew, ymax = ave_mew + sd_mew), width = 0.1) +
  facet_grid(rows = "Location", scales = "free") +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("u, d"^-1)) +
  theme_bw()
```

```{r}
doubling <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y =ave_doubling), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_doubling - sd_doubling, ymax = ave_doubling + sd_doubling), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("Doubling Time, d")) +
  theme_bw()
```


```{r}
delta_bc <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y =ave_delta_bc), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_delta_bc - sd_delta_bc, ymax = ave_delta_bc + sd_delta_bc), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("∆ Bacterial Carbon, umol C L"^-1)) +
  theme_bw()
```


```{r}
delta_bc <- bar.data %>% 
  ggplot(aes(x = factor(Treatment, levels = levels), y =ave_delta_bc), group = interaction(Location, Treatment)) +
  geom_col(color = "black", fill = "white") +
  geom_errorbar(aes(ymin = ave_delta_bc - sd_delta_bc, ymax = ave_delta_bc + sd_delta_bc), width = 0.1) +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  labs(x = "", y = expression("∆ Bacterial Carbon, umol C L"^-1)) +
  theme_bw()
```


```{r fig.height=6, fig.width=14}
delta_bc + mew + doubling + plot_annotation(tag_levels = "A")
```


# Save data

```{r}
saveRDS(averages, "~/Desktop/GITHUB/144l_students/Output_Data/week3/BactAbundanceData.rds")

```

